---
title: "Intermediate R"
author: "Timo Meiendresch"
knit: (function(input_file, encoding) {
  out_dir <- 'html_files';
  rmarkdown::render(input_file,
  encoding=encoding,
  output_file=file.path(dirname(input_file), out_dir, 'Intermediate_R.html'))})
  
---
# Conditionals and Control Flow


## Relational Operators

- Relation of R objects
- Equality (==), Inequality (!=), greater than (>), ...
- In strings: Alphabetical order!
- Note that boolean TRUE coerces to 1 and FALSE to 0

## Logical Operators
Combine several queries (e.g. using several relational operators at once)

- AND operator &
- OR Operator | 
- NOT operator !

```{r}
x <- 12
x > 5 & x < 14
x < 15 | x == 1
```

## Conditional Statements

- if condition 

```r
if (condition) {
  expression
}
```
- if-else condition

```{r}
x <- -3
if(x < 0){
  print("x is a negative number")
} else {
  print("x is not a negative number")
}
```

- else if statement 

```r
if(condition1){
  expr1
} else if(conditon2) {
  expr2
} else {
  expr3
}
```
Extending the example:
```{r}
x <- 0
if(x < 0){
  print("x is a negative number")
} else if(x == 0) {
  print("x is zero")
} else {
  print("x is not a negative number")
}
```

# Loops

### While loop

```r
while(condition){
  expr
}
```

```{r}
ctr <- 1
while(ctr <= 7){
  print(paste("ctr:", ctr))
  ctr <- ctr + 1
}
ctr
```

- break statement

```{r}
ctr <- 1
while (ctr <= 7){
  if(ctr %% 5 == 0){
    break
  }
  print(paste("ctr:", ctr))
  ctr <- ctr +1
}

```

### for loop

```r
for(var in seq){
  expr
}
```
- break: breaks the loop
- next: skips whenever condition is met

```{r}
# for loop, break statement
cities <- list("New York", "Paris","Aachen", "Cologne")
for (city in cities){
  if(nchar(city) == 6){
    break
  }
  print(city)
}

# for loop, next statement
for (city in cities){
  if(nchar(city) == 6){
    next
  }
  print(city)
}
```

- Alternative looping

```{r}
for (i in 1:length(cities)){
  print(paste("City:", i))
  print(cities[i])
}
```

# Functions

- functions work like black box
- argument matching by position or name

```{r}
# help on arguments
args(sd)

# help on function 
#?sd

# linkedin and facebook vectors
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Calculate the mean of the sum
avg_sum <- mean(linkedin + facebook)

# Calculate the trimmed mean of the sum
avg_sum_trimmed <- mean(linkedin + facebook, trim = 0.2)

# Inspect both new variables
avg_sum; avg_sum_trimmed
```

### Writing functions

```r
my_fun <- function(arg1, arg2){
  body
}
```

```{r}
# triple function
triple <- function(x){
  return(3*x)
}
triple(6)

# math_magic()
math_magic <- function(a, b=1){
  a*b + a/b
}
math_magic(4)

```

```{r}
# linkedin and facebook
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# The interpret() can be used inside interpret_all()
interpret <- function(num_views) {
  if (num_views > 15) {
    print("You're popular!")
    return(num_views)
  } else {
    print("Try to be more visible!")
    return(0)
  }
}

# Define the interpret_all() function
interpret_all <- function(views, return_sum=TRUE) {
  count <- 0
  for (v in views) {
  count <- count + interpret(v)
  }
  if (return_sum==TRUE) {
  return(count)
  } else {
  return(NULL)
  }
}

# Call the interpret_all() function
interpret_all(linkedin)
interpret_all(facebook)
```

### Packages

```{r}
# Load the ggplot2 package
library(ggplot2)

# try the qplot() function
qplot(mtcars$wt, mtcars$hp)

# Check out the currently attached packages again
search()
```

# The Apply Family
Instead of loops one can use one of the apply functions

### lapply
```{r}
nyc <- list(pop = 8405837,
            boroughs = c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island"),
            capital = FALSE)

# using a for-loop
for (info in nyc){
  print(class(info))
}

# Use lapply
lapply(nyc, FUN = class)

# cities
lapply(cities, nchar)

# convert to vector
v1 <- unlist(lapply(cities, nchar))
mean(v1)
```

- Note that lapply always returns a list irrespective of input

Next, we'll write a function ourself and use it in connection to lapply

```{r}
# oil 
oil_prices <- round(abs(rnorm(n = 6, mean=2.5, sd=0.5)), 2)
oil_prices

# function 
multiply <- function(x, factor){
  return(factor*x)
}

# both elements together
unlist(lapply(oil_prices, multiply, factor = 5))
```

More examples:

```{r}
# The vector pioneers has already been created for you
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")

# Split names from birth year
split_math <- strsplit(pioneers, split = ":")

# Convert to lowercase strings: split_low
split_low <- lapply(split_math, tolower)

# Take a look at the structure of split_low
str(split_low)
# Write function select_first()
select_first <- function(x) {
  x[1]
}

# Apply select_first() over split_low: names
names <- lapply(split_low, select_first)

# Write function select_second()
select_second <- function(x){
  x[2]
}

# Apply select_second() over split_low: years
years <- lapply(split_low, select_second)
```

### sapply 
- return of a list was necessary for different heterogenous objects in list
- what if all have same type?
- sapply: short for simplified apply

```{r}
# cities again
sapply(cities, nchar)

# 
```

### Recap

- lapply(): apply function over list or vector; output = list
- sapply(): apply function over list or vector; try to simplify list to array

### vapply

- vapply: apply function over list or vector explicitly specify output format

```{r}
# cities, again
sapply(cities, nchar)
vapply(cities, nchar, numeric(1))

# ...


```

# Utilities

```{r}
# some nested functions
li <- list(log = TRUE,
           ch = "hello", 
           int_vec = sort(rep(seq(8,2, by = -2), times =2)))
li
str(li)

# check and convert type
is.list(li)
is.list(c(1,2))
is.list(as.list(c(1,2)))

# append(), rev()
rev(li)
str(append(li, rev(li)))

```

### Regular Expressions
Search for patterns, ...

- grep(), grepl()
- sub(), gsub()

```{r}
animals <- c("cat", "moose", "impala")
grepl(pattern = "a", x=animals)

# starts with ca
grepl(pattern = "^ca", animals)
# ends with a
grepl(pattern = "a$", animals)

# grep returns indices
grep(pattern = "^ca", animals)
grep(pattern = "a$", animals)

# replacement first only
sub(pattern = "a|o|e|i", replacement = "_", x = animals)

# complete replacement
gsub(pattern = "a|o|e|i", replacement = "_", x = animals)

#####################################
# The emails vector has already been defined for you
emails <- c("john.doe@ivyleague.edu", "education@world.gov", "dalai.lama@peace.org",
            "invalid.edu", "quant@bigdatacollege.edu", "cookie.monster@sesame.tv")

# Use grepl() to match for "edu"
grepl("edu", emails)

# Use grep() to match for "edu", save result to hits
hits <- grep("edu", emails)

# Subset emails using hits
emails[hits]

# Use grepl() to match for .edu addresses more robustly
grepl("@.*\\.edu$", emails)

# Use grep() to match for .edu addresses more robustly, save result to hits
hits <- grep("@.*\\.edu$", emails)

# Subset emails using hits
emails[hits]

# Use sub() to convert the email domains to datacamp.edu
sub("@.*\\.edu$", replacement = "@datacamp.edu", emails)
```

### Times & Dates

```{r}
today <- Sys.Date()
today
class(today)

# 
now <- Sys.time()
now
class(now)

# create date objects
my_date <- as.Date("1971-05-14")
my_date
class(my_date)

# 
my_date <- as.Date("1971-14-05", format="%Y-%d-%m")
my_date

# create POSIXct objects
my_time <- as.POSIXct("1971-05-14 11:25:15")
my_time
my_time + 1

my_time2 <- as.POSIXct("1990-11-16 12:00:00" )

# difference
my_time2 - my_time

# both in comparison to 1970
unclass(my_date)
unclass(my_time)

# Definition of character strings representing dates
str1 <- "May 23, '96"
str2 <- "2012-03-15"
str3 <- "30/January/2006"

# Convert the strings to dates: date1, date2, date3
date1 <- as.Date(str1, format = "%b %d, '%y")
date2 <- as.Date(str2, format = "%Y-%m-%d")
date3 <- as.Date(str3, format = "%d/%B/%Y")


# Convert dates to formatted strings
format(date1, "%A")
format(date2, "%d")
format(date3, "%b %Y")

# check
#?strptime

# Definition of character strings representing times
str1 <- "May 23, '96 hours:23 minutes:01 seconds:45"
str2 <- "2012-3-12 14:23:08"

# Convert the strings to POSIXct objects: time1, time2
time1 <- as.POSIXct(str1, format = "%B %d, '%y hours:%H minutes:%M seconds:%S")

time2 <- as.POSIXct(str2, format = "%Y-%m-%d %H:%M:%S")

# Convert times to formatted strings
format(time1, "%M")
format(time2, "%I:%M %p")

```



