---
title: "4. Writing Functions in R"
author: "Timo Meiendresch"
knit: (function(input_file, encoding) {
  out_dir <- 'html_files';
  rmarkdown::render(input_file,
  encoding=encoding,
  output_file=file.path(dirname(input_file), out_dir, 'Intermediate_R.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions in R
Recall the elements of a function in R: 

- formals
- body
- environment

```{r}
# typical function definition
add <- function(x, y=1){
  x+y
}

# 
args(add)
formals(add)
body(add)
environment(add)
```

- Last expression evaluated in a function is the return value
- `return(value)` forces the function to stop execution and return value 
- anonymous functions: functions without a name

```{r}
# anonymous function
(function(x){x+1})(2)
```

Three parts of a function:

- arguments
- body
- environment 

### Environments
Scoping describes how R looks up values by name. If a name isn't defined inside a function, R will look one level up.

```{r}
x <- 2
g <- function(){
  y <- 1
  c(x,y)
}

g()

# 


```

Scoping describes where, not when, to look for a value

```{r}
# depends on environment 
f <- function() x
x <- 15
f()

# x in env is change
x <- 20
f()
```

- When you call a function, a new environment is made for the function to do its work.
- New env is populated with the argument values
- Objects are looked for first in this environment

### Data Structures
```{r}
typeof(letters)
typeof(1:10)
typeof(NULL)
length(NULL)
typeof(NA)
length(NA)
```

- Missing values are contagious!

Lists:

- Useful because they can contain heterogeneous objects
- complicated return objects are lists, ie from lm()

### for loops



