---
title: "Intermediate R"
author: "Timo Meiendresch"
output: html_document
---
# Conditionals and Control Flow


## Relational Operators

- Relation of R objects
- Equality (==), Inequality (!=), greater than (>), ...
- In strings: Alphabetical order!
- Note that boolean TRUE coerces to 1 and FALSE to 0

## Logical Operators
Combine several queries (e.g. using several relational operators at once)

- AND operator &
- OR Operator | 
- NOT operator !

```{r}
x <- 12
x > 5 & x < 14
x < 15 | x == 1
```

## Conditional Statements

- if condition 

```r
if (condition) {
  expression
}
```
- if-else condition

```{r}
x <- -3
if(x < 0){
  print("x is a negative number")
} else {
  print("x is not a negative number")
}
```

- else if statement 

```r
if(condition1){
  expr1
} else if(conditon2) {
  expr2
} else {
  expr3
}
```
Extending the example this leads to:
```{r}
x <- 0
if(x < 0){
  print("x is a negative number")
} else if(x == 0) {
  print("x is zero")
} else {
  print("x is not a negative number")
}
```

# Loops

